Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-26 paper/paper.tex
pass through section 2. added more explanation to the type system.

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Short and Mechanized Logical Relation for Dependent Types}

@@

\begin{figure}[h]
\[
\begin{array}{lcll}
\mathit{Natural\ numbers}\\
[[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\

\mathit{Contexts}\\
[[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ A a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{if} \\ \\
% \mathit{Renaming}\\
% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitution}\\
[[rho]] & \in & [[SNat -> STm]] &
\end{array}
\]
  \caption{Syntax of \lang \scw{Where does the name come from?}\yl{H =
    hierarchy though it needs a better name}}
  \label{fig:syntax}
\end{figure}

In this section, we present the dynamics and statics of the
dependent type theory whose logical consistency will be proven in
Section~\ref{sec:logrelproof}. For conciseness, we refer to this system
as \lang.

The syntax of \lang can be found in Figure~\ref{fig:syntax}.
We use the unscoped de Bruijn representation for both our Coq development and
the informal presentation in our paper since it is more amenable to
automation (Section~\ref{sec:automation}) and leaves no ambiguity
about variable freeness conditions.

%  \scw{Give an overview of the figure}
% \scw{Define a convention that separates your use of natural numbers. Sometimes they
% are universe levels and sometimes they are de Bruijn indices. Better to have separate
% metavariables for each. (And it is possible to generalize universe
% levels to structures other than natural numbers. Ask Jonathan.} \yl{resolved?}
As a dependent type theory, terms and types are collapsed into the same
syntactic category. The type $[[Set i]]$ represent universe
types and $[[var n]]$ represent de Bruijn variables. While
$[[i]],[[j]],$ and $[[n]]$ are all metavariables representing natural
numbers, we always use $[[n]]$ to represent term variables that
participate in substitution and $[[i]],[[j]]$ to represent universe levels.
Dependent functions take the form $[[Pi A B]]$ and
we use the notation $[[A -> B]]$ when the output type $[[B]]$ is not
dependent on the input variable.
Finally, we include in \lang{} the intensional identity type $[[a ~ b : A]]$ whose
proofs can be eliminated by the J-eliminator $[[J c a b p]]$, where
$[[p]]$ is an equality proof between $[[a]]$ and $[[b]]$, and $[[c]]$
is the term whose type is to be casted.
\scw{Example of what you can use identity types for? Or an example of
a program that uses J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \mathit{IdentityTm} \\
        [[idtm n]] & := & [[var n]] \\ \\
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        \mathit{ConsSubst ([[rho .: a]])} \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
        \mathit{Curried\ Add} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]}
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \mathit{Lifting} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \mathit{Substitution ([[ a { rho }  ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

The \lang language is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ Bool if var 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

We adapt from \citet{autosubst2} the notations for simultaneous
renaming, substitution, and other auxiliary definitions used in
our paper, summarized in Figure~\ref{fig:auxdef}. We need those
definitions to specify the reduction relation and the typing relation
for \lang{}. The metavariable $[[rho]]$ represents substitutions,
mappings from de Bruijn indices to terms.
One example of a substitution is
the function $\ottkw{id}_{tm}$, which takes a de Bruijn index directly
injects it as a term variable. The $\uparrow^{[[n]]}$ operator is
more general and adds $[[n]]$ to the index before injecting it as a
term. The $[[(rho .: a)]]$ operation allows us to extend a substitution
$[[rho]]$ with an extra element $[[a]]$; the extended substitution maps the variable
$[[var 0]]$ to $[[a]]$ and variables $[[var Suc i]]$ to $[[rho i]]$.

The substitution operator, which takes the form $[[a {rho}]]$,
traverses the syntax of $[[a]]$ and replaces each variable $[[var i]]$
with the term $[[rho i]]$. When traversing under binders (e.g. in the
$[[(\ a) { rho }]]$ case), the $\Uparrow$ operator prevents $[[var
0]]$, the bound variable, from being replaced by
$[[rho]]$. From the definition of $[[up rho]]$, we see that it keeps
$[[var 0]]$ unchanged during substitution and replaces each $[[var
Suc i]]$ with $[[rho i {up 1}]]$. The extra shifting with $[[up 1]]$
is required so the variables in terms from the codomain of $[[rho]]$ skip over
the newly introduced binder to refer to the correct binding
location. The substitution operator is referred to as simultaneous
substitution as it substitutes all variables at once. It is possible
to recover single substitution by composing the extension operator and
the identity substitution: $[[a { b }]] := [[a { idtm .: b }]]$. Since
the definition of semantic typing in Section~\ref{sec:logrelproof}
relies on simultaneous substitution, we find it more convenient to
formulate simultaneous substitution directly rather than recovering it
from single substitution.

@@

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we need to first specify its
untyped equational theory. We choose to represent the untyped
equational theory in an algorithmic style using the notion of parallel
reduction, defined in Figure~\ref{fig:par} (congruence rules are omitted).
 Parallel reduction takes
the form $[[a => b]]$ and we denote its transitive as $[[a =>+
b]]$. We say that two terms $[[a0]]$ and $[[a1]]$ are convertible,
denoted by the notation $[[a0 <=> a1]]$, if there exists some term
$[[b]]$ such that $[[a0 =>+ b]]$ and $[[a1 =>+ b]]$. The symmetric
notation of convertibility suggests that it is an equivalence
relation. We sketch out the sequence of lemmas required to derive this
property below and omit the details. Our technique for proving the
properties about parallel reduction is based on
\citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.
\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
\ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
equivalence relation. It is then trivial to show that all untyped
$\beta$-equivalence rules are admissible, though we omit this proof
from our mechanization and will from now on focus on parallel
reduction since the diamond property (Lemma~\ref{lemma:pardiamond})
makes it easier to work with. A detailed argument of the equivalence
between $[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}.

Untyped equality allows us to show the
equivalence between our more structured notion of convertibility
 ($[[a <=> b]]$) and the declarative, untyped $\beta$-equivalence
independently from the type system and the logical relation through
well-established syntactic approaches. The directedness of the
convertibility relation will be leveraged in
Section~\ref{sec:logreldep} to justify the conversion rule of our
typing judgment, which we will introduce shortly.

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Refl, Eq, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

In \rref{T-Var}, $[[G n]]$ is the partial function defined
by the equations $[[ (G ++ A)  0 ]] := [[A]]$ and $[[ (G ++ A) Suc
n ]] := [[G n]]$.
The precondition $[[n < | G | ]]$, where $[[ | G |
]]$ represents the length of the context $[[G]]$, ensures that $[[G
n]]$ has a defined output. The type $[[G n]]$ needs to be shifted
by $[[Suc n]]$ since the types stored in the context $[[G]]$ are
scoped differently. In the typing judgment $[[G |- a : A]]$, we want
both $[[a]]$ and $[[A]]$ to have the same scope. However, the closer a
type is to the head of $[[G]]$, the smaller de Bruijn it has since
there is less variables in the context it can refer to (see \rref{Ctx-Cons}). Shifting $[[G
n]]$ by $[[Suc n]]$ weakens the type so it is scoped consistently with
the full context $[[G]]$ rather than a truncated version of
$[[G]]$. In \rref{T-J}, the elimination form for the identity type,
the terms $[[a]]$ and $[[A]]$ need to be shifted in the context for a similar
reason; an extra type $[[A]]$ appears after the $[[G]]$ and therefore
the terms $[[a]]$ and $[[A]]$ in the identity type both need to be shifted
to refer to the same binders.

\Rref{T-Conv} uses the convertibility relation from earlier
as our equality judgment for type conversion. The use of an untyped
relation for type conversion is reminiscent of Barendregt's Pure Type
Systems~\citet{barendregt1991introduction} and makes our formulation different
from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental equality
takes the form $\Gamma \vdash a \equiv b : A$, from which one can
usually derive  $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving
subject reduction.

\citet{siles2012pure} shows the equivalence of Barendregt's
Pure Type System~\citep{barendregt1991introduction}, which employs
untyped equality, and its variant that uses typed judgmental
equality. This assures us that we do not lose much generality working
with a system with untyped conversion.
We include a detailed discussion in Section~\ref{sec:discuss} on how
type-directed and untyped equality affect the applicability of our
technique.

@@

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{ l |  l  | l | l | l }
      & \header{Universes} & \header{Inductives} & \header{Conversion}
      & \header{Large Elim} \\
    \hline
    \lang{} & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗}\\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} \\
    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
    % \hline
    CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}
% \end{minipage}

  \begin{tabular}{l}
  \\
    $\lambda^\theta$ (logical fragment only) \citet{casinghino:combining-proofs-programs} \\
    NBE-in-Coq \citet{nbeincoq} \\
    $\lambda^{\Pi U\mathbb{N}}$ \citet{decagda} \\
    MLTT-a-la-Coq \citet{martin-lof-a-la-coq} \\
    Core Nuprl \citet{anand2014towards} \\
    CIC \citet{barras2012semantical} \\
  \\
  Universes: Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
  Conversion: Typed (T), Untyped (U), Extensional (E) \\
  \end{tabular}

>>>>

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}

@@

\begin{figure}[h]
\[
\begin{array}{lcll}
% \mathit{Natural\ numbers}\\
% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[var n]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi A B]]\ |\ [[\ a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{conditional expression} \\ \\
% \mathit{Renaming}\\
% [[xi]] & \in & [[SNat -> SNat]] & \\ \\
\mathit{Substitutions}\\
[[rho]] & \in & [[SNat -> STm]] & \\ \\ 
\mathit{Typing Contexts}\\
 [[G]]       & ::= & [[empty]]\ |\ [[G ++ A]] &  \\ \\
\end{array}
\]
  \caption{Syntax of \lang \scw{Where does the name come from?}\yl{H =
    hierarchy though it needs a better name}}
  \label{fig:syntax}
\end{figure}

In this section, we present the dependent type theory whose logical
consistency will be proven in Section~\ref{sec:logrelproof}. For conciseness,
we refer to this system as \lang.

The syntax of \lang can be found in Figure~\ref{fig:syntax}.  As a dependent
type theory, terms and types are collapsed into the same syntactic
category. The type $[[Set i]]$ represent universe types and $[[var n]]$
represent de Bruijn variables~\cite{debruijn}.  While $[[i]],[[j]],$ and
$[[n]]$ are all metavariables representing natural numbers, we always use
$[[i]],[[j]]$ to represent universe levels and $[[n]]$ to represent term
variables.  We use de Bruijn indices for both our Coq development and the
informal presentation here because it is more amenable to automation
(Section~\ref{sec:automation}) and leaves no ambiguity about variable freeness
conditions. As a result, variables do not appear in binding forms, such as in
abstractions $[[ \ a ]]$ and dependent function types $[[Pi A B]]$.  This latter
form binds a new variable of type $[[A]]$ that can appear in the codomain of
the type $[[B]]$.  We use the notation $[[A -> B]]$ when the output type
$[[B]]$ is not dependent on the input variable.
We include in \lang{} the intensional identity type
$[[a ~ b : A]]$ whose proofs can be eliminated by the J-eliminator
$[[J c a b p]]$, where $[[p]]$ is an equality proof between $[[a]]$ and
$[[b]]$, and $[[c]]$ is the term whose type is to be casted.
Finally, \lang{} includes booleans, with standard syntax.
\scw{Example of
 what you can use identity types for? Or an example of a program that uses
J?}

@@

% \scw{Need to explain the notations in the text. What are these operations?
% What do readers need to understand about them? Help me understand
% the figure} \yl{resolved}
% \scw{As this is a tutorial paper, you'll need to explain more about how de Bruijn
% indices work.} \yl{resolved?}
% We omit most of the
% definitions of renaming and substitution and only show the definition
% of a few representative cases of substitution.
\begin{figure}[ht]
  \begin{equation*}
    \begin{split}
      \begin{array}{lll}
        % \mathit{IdentityRen} \\
        % [[id i]] & := & [[i]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Identity substitution: }} \mathsf{id}_{\mathit{tm}}} \\
        [[idtm n]] & := & [[var n]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Weakening substitution: }} \uparrow^{n}} \\
        [[up n m]] & := & \ottkw{v}_{[[n]] + [[m]]} \\ \\
        %
        % \mathit{ConsRen ([[xi .: n]])} \\
        % ([[xi .: n]])([[0]]) & := & [[n]] \\
        % ([[xi .: n]])([[Suc i]]) & := & [[xi i]] \\ \\
        %
        \multicolumn{3}{l}{\mbox{\emph{Extension operation: }} ([[rho .: a]]) } \\
        ([[rho .: a]])([[0]]) & := & [[a]] \\
        ([[rho .: a]])([[Suc n]]) & := & [[rho n]] \\ \\
      \end{array}
    \end{split}
    \qquad \qquad
    \begin{split}
      \begin{array}{lll}
        \multicolumn{3}{l}{\mbox{\emph{Lifting operation: }} \Uparrow\rho} \\
        [[( up rho ) 0]] & := & [[var 0]] \\
        [[( up rho ) Suc n]] & := & [[ rho n < up 1 > ]] \\ \\
        \multicolumn{3}{l}{\mbox{\emph{Substitution operation:}} ([[ a { rho } ]])} \\
        [[var n { rho }  ]] & := &  [[rho n]] \\
        [[(Pi A B) { rho }]] & := & [[Pi A { rho } B { up rho }]] \\
        [[(a b) { rho }]] & := & [[a { rho } (# b { rho } #)]] \\
        [[(\ a) { rho }]] & := & [[\ (a { up rho })]] \\
        \ldots
      \end{array}
    \end{split}
  \end{equation*}
  \caption{Auxiliary Functions over Syntax}
  \label{fig:auxdef}
\end{figure}

@@

The \lang language is expressive enough to support large
eliminations, the ability to compute a type using a term as input. For
example, the function $[[\ if var 0 Bool Bool -> Bool]]$ returns
either $[[Bool]]$ or $[[Bool -> Bool]]$ depending on whether the input
is $[[true]]$ or $[[false]]$.

We adapt from \citet{autosubst2} the notations for simultaneous renaming,
substitution, and other auxiliary definitions used in our paper, summarized in
Figure~\ref{fig:auxdef}. We need these definitions to specify the reduction
and typing relations for \lang{}. The metavariable $[[rho]]$ represents
substitutions, which are mappings from de Bruijn indices to terms.  One
example of a substitution is the function $\ottkw{id}_{tm}$, the identity
substitution that maps each index to the corresponding term variable.  The
$\uparrow^{[[n]]}$ substitution is more general and adds $[[n]]$ to the index
before injecting it as a term. The extension operation, $[[(rho .: a)]]$,
allows us to extend a substitution $[[rho]]$ with an extra element $[[a]]$;
the result maps the variable $[[var 0]]$ to $[[a]]$ and variables
$[[var Suc i]]$ to $[[rho i]]$.

The substitution operator, which takes the form $[[a {rho}]]$,
traverses the syntax of $[[a]]$ and replaces each variable $[[var i]]$
with the term $[[rho i]]$. When traversing under binders (e.g. in the
$[[(\ a) { rho }]]$ case), the $\Uparrow$ operator prevents $[[var
0]]$, the bound variable, from being replaced by
$[[rho]]$. From the definition of $[[up rho]]$, we see that it keeps
$[[var 0]]$ unchanged during substitution and replaces each $[[var
Suc i]]$ with $[[rho i {up 1}]]$. The shifting with $[[up 1]]$
is required so the variables in terms from the codomain of $[[rho]]$ skip over
the newly introduced binder to refer to the correct binding
location.

The substitution operator is referred to as simultaneous substitution as it
substitutes for all variables at once. It is possible to recover single
substitution by composing the extension operator and the identity
substitution: $[[a { b }]] := [[a { idtm .: b }]]$. In our proofs, we find it
more convenient to formulate simultaneous substitution directly rather than
recovering it from single substitution.  In particular, this shows up in the
definition of semantic typing in Section~\ref{sec:logrelproof}, which relies
on simultaneous substitution.

\subsection{Definitional equality via parallel reduction}

@@

% \scw{motivate parallel reduction. Why are you telling us this now. Is this
%   an explanation of definitional equality and its properties?}
Before we specify the typing rules, we first specify the equational theory
used in its conversion rule (\rref{T-Conv} in Figure~\ref{fig:typing}). This
relation is often called ``definitional equality'' in dependent type theories
because it defines the equivalence that the syntactic type system works
``up-to''.

For \lang{}, we use an untyped equational theory, defined in an algorithmic
style using the notion of parallel reduction, written $[[a => b]]$, and shown
in Figure~\ref{fig:par}. (For brevity, reflexivity and congruence rules are
omitted from this figure). The notation $[[a =>+ b]]$ indicates the transitive
closure of parallel reduction.

\begin{definition}[Convertibility]
  Two terms $[[a0]]$ and $[[a1]]$ are \emph{convertible}, written
  $[[a0 <=> a1]]$, if there exists some term $[[b]]$ such that $[[a0 =>+ b]]$
  and $[[a1 =>+ b]]$.
\end{definition}

Convertibility is an equivalence relation. We sketch the sequence of lemmas
required to derive this property below and omit the details. The key step in
proving transitivity, is showing that parallel reduction satisfies the diamond
property.  Our technique for proving the properties about parallel reduction
is based on \citet{takahashi-parallel-reduction}. A modern exposition of the
same technique can be found in \citet{plfa22.08}.

\begin{lemma}[Par Refl\footnote{\dotv{join.v}{Par\_refl}}]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}
\begin{lemma}[Par cong\footnote{\dotv{join.v}{par\_cong}}]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{corollary}[Par subst\footnote{\dotv{join.v}{par\_subst}}]
  \label{lemma:parsubst}
  If $[[a0 => a1]]$, then $[[a0 {b} => a1 {b}]]$ for arbitrary $[[b]]$.
\end{corollary}
\begin{lemma}[Par diamond\footnote{\dotv{join.v}{par\_confluent}}]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
\begin{lemma}[Convertibility refl\footnote{\dotv{join.v}{Coherent\_reflexive}}]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility sym\footnote{\dotv{join.v}{Coherent\_symmetric}}]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}
\begin{lemma}[Convertibility trans\footnote{\dotv{join.v}{Coherent\_transitive}}]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
% From Lemmas~\ref{lemma:coherencerefl}, \ref{lemma:coherencesym}, and
% \ref{lemma:coherencetrans}, we conclude that convertibility is indeed an
% equivalence relation.

Our definition of convertibility for \lang{} is unusual in that it is directly
defined via parallel reduction, instead of using the related notion of
$\beta$-equivalence~\cite{barendregt:pts,cic}. This choice does not change the
language definition; a detailed argument of the equivalence between
$[[a <=> b]]$ and untyped $\beta$-equivalence can be found in
\citet{barendregt:lambda-calculi-with-types} and
\citet{takahashi-parallel-reduction}. However, this choice simplifies later
proofs, as we discuss in Section~\ref{XXX}. \scw{Forward reference to further
  discussion, comparing with whnf reduction?}

Our definition of equality is untyped: the judgement does not require the two
terms to type check and have the same type. The use of an untyped relation for
type conversion is reminiscent of Barendregt's Pure Type
Systems~\citet{barendregt1991introduction} and makes our formulation different
from languages such as MLTT~\citep{Martin-Lof-1973}, where the judgmental
equality takes the form $\Gamma \vdash a \equiv b : A$, from which one can
usually derive $[[G |- a : A]]$ and $[[G |- b : A]]$ after proving subject
reduction. By working with an untyped judgement, we can establish its
properties independently from the type system and the logical relation, using
well-established syntactic approaches.

\citet{siles2012pure} shows the equivalence of Barendregt's Pure Type
System~\citep{barendregt1991introduction}, which employs untyped equality, and
its variant that uses typed judgmental equality. This assures us that we do
not lose generality working with a system with untyped
conversion. Furthermore, discuss how this definition can be extended with
$\eta$-equivalence of functions in Section~\ref{XXX}. Finally, we compare this
definition with type-directed approaches to equality in
Section~\ref{sec:discuss}.

\subsection{Syntactic Typing}

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{Empty, Cons}
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
  Eq, Refl, J, If, Bool, True, False, Void}
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

In \rref{T-Var}, $[[G n]]$ is the partial function defined by the equations
$[[ (G ++ A) 0 ]] := [[A]]$ and $[[ (G ++ A) Suc n ]] := [[G n]]$.  The
precondition $[[n < | G | ]]$, where $[[ | G | ]]$ represents the length of
the context $[[G]]$, ensures that $[[G n]]$ has a defined output. The type
$[[G n]]$ needs to be shifted by $[[Suc n]]$ since the types stored in the
context $[[G]]$ are scoped differently. In the typing judgment
$[[G |- a : A]]$, we want both $[[a]]$ and $[[A]]$ to have the same
scope. However, the closer a type is to the head of $[[G]]$, the smaller de
Bruijn it has since there is less variables in the context it can refer to
(see \rref{Ctx-Cons}). Shifting $[[G n]]$ by $[[Suc n]]$ weakens the type so
it is scoped consistently with the full context $[[G]]$ rather than a
truncated version of $[[G]]$.

Although the use of de Bruijn indices complicates the variable rule, the rules
that involve binding, such as \rref{T-Pi,T-Abs} are straightforward.

\Rref{T-Conv} uses the convertibility relation from earlier
as our equality judgment for type conversion.

The elimination form for booleans, \rref{T-If} demonstrates dependent pattern
matching.  The result type of this expression $[[A {a}]]$, is composed of some
\emph{motive} $[[A]]$, a type where its single free variable has been replaced
with the condition of the if expression. When typing the true branch, this
substitution replaces the variable by $[[true]]$, and similar for the false
branch. As a result, the type system communicates the information gained from
the run-time test to each of the branches of the expression.

A similar sort of dependent pattern matching occurs when eliminating identity
types. Such types are checked for well-formedness with \rref{T-Eq} and
introduced by \rref{T-Refl}. In \rref{T-J}, the elimination form, the subterm
$p$ is a proof of an equality between $[[a]]$ and $[[b]]$. The subterm $[[c]]$
is the body of the elimination form. In this rule, $[[B]]$ is called a
\emph{motive} and has two free variables.  The substitution for these
variables changes from $[[b]]$ to $[[a]]$ and from $[[p]]$ to $[[refl]]$,
witnessing the information gained through dependent pattern matching.  In the
typing rule, the terms $[[a]]$ and $[[A]]$ need to be shifted in the context
for a similar reason; an extra type $[[A]]$ appears after the $[[G]]$ and
therefore the terms $[[a]]$ and $[[A]]$ in the identity type both need to be
shifted to refer to the same binders.

@@

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{ l |  l  | l | l | l }
      & \header{Universes} & \header{Inductives} & \header{Conversion}
      & \header{Large Elim} \\
    \hline
    \lang{} & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗}\\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} \\
    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} \\
    \hline
    CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    CoreSpec & \\
    PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}
% \end{minipage}

  \begin{tabular}{ll}
  \\
    $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
    NBE-in-Coq& \citet{nbeincoq} \\
    $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
    MLTT-a-la-Coq &\citet{martin-lof-a-la-coq} \\
    Core Nuprl &\citet{anand2014towards} \\
    CIC &\citet{barras2012semantical} \\
    CoreSpec & \\
    PCUIC & \\

  \\
  Universes: &Countable ($\mathbb{N}$), Zero (0), One (1), Impredicative ($\mathbb{I}$) \\
  Conversion:& Typed (T), Untyped (U), Extensional (E) \\
  \end{tabular}
